---
title: "R Notebook"
output: html_notebook
---

# Load Libraries
```{r}
# Load necessary libraries
library(tidyverse)   # For data manipulation and visualization (includes dplyr and ggplot2)
library(dplyr)
library(lubridate)   # For date-time manipulation
library(readxl)      # For reading Excel files
library(naniar)
library(stringr)
```
# Data Preparation and Cleaning
```{r}
# Read data from Excel sheets into separate data frames
file_path <- "StatIIInterviewDataset_Aug 2024.xlsx"  # Path to the Excel file
incident_times <- read_xlsx(file_path, sheet = "Incident Times")  # Read "Incident Times" sheet
cardiac_arrest <- read_xlsx(file_path, sheet = "Cardiac Arrest")  # Read "Cardiac Arrest" sheet
vitals <- read_xlsx(file_path, sheet = "Vitals")  # Read "Vitals" sheet
patient <- read_xlsx(file_path, sheet = "Patient")  # Read "Patient" sheet
response <- read_xlsx(file_path, sheet = "Response")  # Read "Response" sheet

# Print column names for each data frame to check the structure
colnames(incident_times)  # Column names of incident_times
colnames(cardiac_arrest)  # Column names of cardiac_arrest
colnames(vitals)          # Column names of vitals
colnames(patient)         # Column names of patient
colnames(response)        # Column names of response

# Print structure of each data frame to understand the data types and structure
str(incident_times)
str(cardiac_arrest)
str(vitals)
str(patient) 
str(response)
```
## Consistency checks
```{r}
# Use gsub to remove newline characters from column names
# This makes column names cleaner and easier to work with
colnames(incident_times) <- gsub("\r\n", " ", colnames(incident_times))
colnames(cardiac_arrest) <- gsub("\r\n", " ", colnames(cardiac_arrest))
colnames(vitals) <- gsub("\r\n", " ", colnames(vitals))
colnames(patient) <- gsub("\r\n", " ", colnames(patient))
colnames(response) <- gsub("\r\n", " ", colnames(response))

# Print column names again to check changes
colnames(incident_times)
colnames(cardiac_arrest)
colnames(vitals)
colnames(patient)
colnames(response)
```
## Date-Time as POSIXct
```{r}
# Convert Excel serial date-time to R Date-Time
# Function to convert Excel serial date-time numbers to POSIXct date-time objects
convert_excel_datetime <- function(serial_date) {
  # Base date for Excel
  base_date <- as.POSIXct("1899-12-30", tz = "UTC") # Excel base date
  
  # Convert serial date to POSIXct
  posix_datetime <- base_date + as.numeric(serial_date) * 86400 # 86400 seconds in a day
  
  return(posix_datetime)
}

# Example value to demonstrate conversion
serial_date <- 44592.508437500001  # Example Excel serial date-time number
converted_datetime <- convert_excel_datetime(serial_date)  # Convert to POSIXct
print(converted_datetime)  # Print converted date-time
```

## Identify missing data
```{r}
# Check for missing values in each dataset
# This helps to identify any data completeness issues
sapply(list(incident_times, cardiac_arrest, vitals, patient, response), function(x) sum(is.na(x)))
```
## Update Column Data Types
```{r}
# Update all eTimes variables using the conversion function
# Apply convert_excel_datetime to all columns containing "eTimes" in their names
vitals <- vitals %>%
  mutate(across(contains("eTimes"), ~ convert_excel_datetime(.x)))

# Specifically update the "Vitals Signs Take Date Time (eVitals.01)" column
vitals <- vitals %>%
  mutate(`Vitals Signs Take Date Time (eVitals.01)` = convert_excel_datetime(`Vitals Signs Take Date Time (eVitals.01)`))
```

## Aggregate Vitals Data
```{r}
# Ensure numeric fields are properly formatted
# Convert columns containing "eVitals" to numeric type
vitals <- vitals %>%
  mutate(
    across(
      c(
        'Vitals Systolic Blood Pressure SBP (eVitals.06)', 
        'Vitals Heart Rate (eVitals.10)', 
        'Vitals Respiratory Rate (eVitals.14)'
      ), 
      as.numeric  # Convert specified columns to numeric
    )
  )

# Assuming `vitals_data` is your dataframe containing the vital signs

aggregated_vitals <- vitals %>%
  group_by(`Patient Care Report Number - PCR (eRecord.01)`) %>%
  summarise(across(contains("eVitals") & !contains("eVitals.01"), 
                   list(mean = ~ mean(.x, na.rm = TRUE),
                        min = ~ min(.x, na.rm = TRUE),
                        max = ~ max(.x, na.rm = TRUE))), 
            .groups = "drop",
            `Vitals Signs Take Date Time (eVitals.01)` = first(`Vitals Signs Take Date Time (eVitals.01)`))

```

# Joining the Tables
```{r}
# Joining tables on PCR number to create a combined dataset
# Each table is joined based on the common column "Patient Care Report Number - PCR (eRecord.01)"
data <- incident_times %>%
    left_join(cardiac_arrest, by = c("Patient Care Report Number - PCR (eRecord.01)" = "Patient Care Report Number - PCR (eRecord.01)")) %>%
    left_join(aggregated_vitals, by = c("Patient Care Report Number - PCR (eRecord.01)" = "Patient Care Report Number - PCR (eRecord.01)")) %>%
    left_join(patient, by = c("Patient Care Report Number - PCR (eRecord.01)" = "Patient Care Report Number - PCR (eRecord.01)")) %>%
    left_join(response, by = c("Patient Care Report Number - PCR (eRecord.01)" = "Patient Care Report Number - PCR (eRecord.01)"))
```

```{r}
# Update all eTimes variables using the conversion function
# Apply convert_excel_datetime to all columns containing "eTimes" in their names
data <- data %>%
  mutate(across(contains("eTimes"), ~ convert_excel_datetime(.x)))

# Specifically update the "Vitals Signs Take Date Time (eVitals.01)" column
data <- data %>%
  mutate(`Vitals Signs Take Date Time (eVitals.01)` = convert_excel_datetime(`Vitals Signs Take Date Time (eVitals.01)`))
```

## Identify duplicates
```{r}
combined_data <- data
# Check for duplicate PCR numbers across all datasets
duplicate_pcr <- combined_data %>%
  group_by(`Patient Care Report Number - PCR (eRecord.01)`) %>%
  filter(n() > 1) %>%
  summarise(Count = n())

# View duplicate PCR numbers
print(duplicate_pcr)
```
## Create Age Variable
```{r}
# Combine age variables into a single age variable in years
# Ensure age and units columns are treated as characters for easier handling
combined_data <- combined_data %>%
  mutate(
    `Patient Age (ePatient.15)` = as.numeric(`Patient Age (ePatient.15)`),
    `Patient Age Units (ePatient.16)` = as.character(`Patient Age Units (ePatient.16)`)
  )

# Define the function to convert age to years
convert_age_to_years <- function(age, units) {
  # Handle missing values
  if (is.na(age) || is.na(units)) {
    return(NA)
  }
  
  # Convert units to lower case for consistency
  units <- tolower(units)
  
  # Convert age based on units
  if (units == "months") {
    return(age / 12)  # Convert months to years
  } else if (units == "days") {
    return(age / 365.25)  # Convert days to years (approximate)
  } else if (units == "years") {
    return(age)  # Already in years
  } else {
    # Print a message if an unknown unit is encountered
    warning(paste("Unknown age unit:", units))
    return(NA)
  }
}

# Apply the conversion function to each row
combined_data <- combined_data %>%
  mutate(
    Patient_Age_Years = mapply(convert_age_to_years, `Patient Age (ePatient.15)`, `Patient Age Units (ePatient.16)`)
  )

# Check the result
head(combined_data %>% select(`Patient Age (ePatient.15)`, `Patient Age Units (ePatient.16)`, Patient_Age_Years))
```

# Missing Data
## Visualize Missing Data
```{r}
# Visualize missing data patterns
gg_miss_var(combined_data)  # Visualize missing data by variable
```

## Summarize Missing Data
```{r}
# Summarize missing data
summary(combined_data)

# Proportion of missing data per column
sapply(combined_data, function(x) mean(is.na(x)))
```

# Detecting Outliers
## Using Boxplots
```{r}
# Visualize outliers for vital signs
combined_data %>%
  select(
    `Vitals Systolic Blood Pressure SBP (eVitals.06)_mean`,
    `Vitals Heart Rate (eVitals.10)_mean`,
    `Vitals Respiratory Rate (eVitals.14)_mean`
  ) %>%
  gather(key = "variable", value = "value") %>%
  mutate(variable = str_replace_all(variable, c("_mean$" = "", "\\s*\\(eVitals\\.\\d{2}\\)" = ""))) %>%  # Remove "_mean" and "(eVitals.xx)"
  ggplot(aes(x = variable, y = value)) +
  geom_boxplot() +
  theme_minimal() +
  labs(title = "Boxplot for Vital Signs to Detect Outliers", x = "Variable", y = "Value")

```

# Who is impacted?
## Age
```{r}
# Create age bins and categorize each age
combined_data <- combined_data %>%
  mutate(
    Age_Group = case_when(
      Patient_Age_Years < 18 ~ "Younger than 18",
      Patient_Age_Years >= 18 & Patient_Age_Years <= 24 ~ "18–24",
      Patient_Age_Years >= 25 & Patient_Age_Years <= 44 ~ "25–44",
      Patient_Age_Years >= 45 & Patient_Age_Years <= 64 ~ "45–64",
      Patient_Age_Years >= 65 & Patient_Age_Years <= 74 ~ "65–74",
      Patient_Age_Years >= 75 ~ "75 and older",
      TRUE ~ NA_character_  # Handle any NA values
    )
  )

age_summary <- combined_data %>%
  group_by(Age_Group) %>%
  filter(`Cardiac Arrest During EMS Event (eArrest.01)` %in% c(
    "Yes, Prior to Any EMS Arrival\r\n(includes Transport EMS &\r\nMedical First Responders)",
    "Yes, After Any EMS Arrival\r\n(includes Transport EMS &\r\nMedical First Responders)"
  )) %>%
  summarise(
    count = n()
  ) %>%
  arrange(desc(count))

# Print age summary
print(age_summary)
```
## Gender
```{r}
# Summarize data by Gender
gender_summary <- combined_data %>%
  group_by(`Patient Gender (ePatient.13)`) %>%
  filter(`Cardiac Arrest During EMS Event (eArrest.01)` %in% c(
    "Yes, Prior to Any EMS Arrival\r\n(includes Transport EMS &\r\nMedical First Responders)",
    "Yes, After Any EMS Arrival\r\n(includes Transport EMS &\r\nMedical First Responders)"
  )) %>%
  summarise(
    count = n()
  ) %>%
  arrange(desc(count))

# Print gender summary
print(gender_summary)
```

## Race/Ethnicity
```{r}
# Summarize data by Race/Ethnicity
race_summary <- combined_data %>%
  group_by(`Patient Race LIST (ePatient.14)`) %>%
  filter(`Cardiac Arrest During EMS Event (eArrest.01)` %in% c(
    "Yes, Prior to Any EMS Arrival\r\n(includes Transport EMS &\r\nMedical First Responders)",
    "Yes, After Any EMS Arrival\r\n(includes Transport EMS &\r\nMedical First Responders)"
  )) %>%
  summarise(
    count = n()
  ) %>%
  arrange(desc(count))

# Print race/ethnicity summary
print(race_summary)
```


```{r}
# Define a function to clean and standardize race/ethnicity categories
clean_race_categories <- function(race) {
  race <- tolower(race)  # Convert to lowercase for consistency
  
  # Define broad categories
  if (grepl("white", race)) {
    if (grepl("black", race)) {
      return("White and Black or African American")
    } else if (grepl("hispanic", race)) {
      return("White and Hispanic or Latino")
    } else if (grepl("asian", race)) {
      return("White and Asian")
    } else if (grepl("native hawaiian", race) || grepl("pacific islander", race)) {
      return("White and Native Hawaiian or Other Pacific Islander")
    } else if (grepl("american indian", race)) {
      return("White and American Indian or Alaska Native")
    } else {
      return("White")
    }
  } else if (grepl("black", race)) {
    if (grepl("hispanic", race)) {
      return("Black or African American and Hispanic or Latino")
    } else if (grepl("asian", race)) {
      return("Black or African American and Asian")
    } else if (grepl("native hawaiian", race) || grepl("pacific islander", race)) {
      return("Black or African American and Native Hawaiian or Other Pacific Islander")
    } else if (grepl("american indian", race)) {
      return("Black or African American and American Indian or Alaska Native")
    } else {
      return("Black or African American")
    }
  } else if (grepl("hispanic", race)) {
    if (grepl("asian", race)) {
      return("Hispanic or Latino and Asian")
    } else if (grepl("native hawaiian", race) || grepl("pacific islander", race)) {
      return("Hispanic or Latino and Native Hawaiian or Other Pacific Islander")
    } else if (grepl("american indian", race)) {
      return("Hispanic or Latino and American Indian or Alaska Native")
    } else {
      return("Hispanic or Latino")
    }
  } else if (grepl("asian", race)) {
    if (grepl("native hawaiian", race) || grepl("pacific islander", race)) {
      return("Asian and Native Hawaiian or Other Pacific Islander")
    } else if (grepl("american indian", race)) {
      return("Asian and American Indian or Alaska Native")
    } else {
      return("Asian")
    }
  } else if (grepl("native hawaiian", race) || grepl("pacific islander", race)) {
    if (grepl("american indian", race)) {
      return("Native Hawaiian or Other Pacific Islander and American Indian or Alaska Native")
    } else {
      return("Native Hawaiian or Other Pacific Islander")
    }
  } else if (grepl("american indian", race)) {
    return("American Indian or Alaska Native")
  } else if (grepl("not recorded", race) || grepl("not applicable", race)) {
    return("Not Recorded/Applicable")
  } else {
    return("Other")
  }
}

# Apply the cleaning function to the race data
combined_data <- combined_data %>%
  mutate(
    Cleaned_Race = sapply(`Patient Race LIST (ePatient.14)`, clean_race_categories)
  )
```

```{r}
# Summarize by cleaned race categories
race_summary_cleaned <- combined_data %>%
  group_by(Cleaned_Race) %>%
  filter(`Cardiac Arrest During EMS Event (eArrest.01)` %in% c(
    "Yes, Prior to Any EMS Arrival\r\n(includes Transport EMS &\r\nMedical First Responders)",
    "Yes, After Any EMS Arrival\r\n(includes Transport EMS &\r\nMedical First Responders)"
  )) %>%
  summarise(
    count = n()
  ) %>%
  arrange(desc(count))

# Print the cleaned race/ethnicity summary
print(race_summary_cleaned)
```
# What care is provided?
```{r}
# Summarize types of interventions provided
# View unique values for 'Cardiac Arrest During EMS Event (eArrest.01)' to confirm the expected values
unique(combined_data$'Cardiac Arrest During EMS Event (eArrest.01)')
unique(combined_data$'Cardiac Arrest Etiology (eArrest.02)')
unique(combined_data$'Resuscitation Attempted by EMS LIST (eArrest.03)')
```

```{r}
# Summarize types of interventions provided
interventions_summary <- combined_data %>%
  # Filter to include only rows where cardiac arrest occurred during the EMS event
  filter(`Cardiac Arrest During EMS Event (eArrest.01)` %in% c(
    "Yes, Prior to Any EMS Arrival\r\n(includes Transport EMS &\r\nMedical First Responders)",
    "Yes, After Any EMS Arrival\r\n(includes Transport EMS &\r\nMedical First Responders)"
  )) %>%
  summarise(
    total_cases = n(),  # Total number of cardiac arrest cases during EMS event
    cpr_provided = sum(`Resuscitation Attempted by EMS LIST (eArrest.03)` %in% c("Attempted Defibrillation,\r\nInitiated Chest Compressions,\r\nAttempted Ventilation",
                                                                                 "Attempted Defibrillation,\r\nAttempted Ventilation, Initiated\r\nChest Compressions",
                                                                                 "Initiated Chest Compressions,\r\nAttempted Ventilation",
                                                                                 "Initiated Chest Compressions"), na.rm = TRUE),
    defibrillator_used = sum(`Resuscitation Attempted by EMS LIST (eArrest.03)` %in% c("Attempted Defibrillation,\r\nInitiated Chest Compressions,\r\nAttempted Ventilation",
                                                                                       "Attempted Defibrillation,\r\nAttempted Ventilation, Initiated\r\nChest Compressions"), na.rm = TRUE),
    not_attempted = sum(`Resuscitation Attempted by EMS LIST (eArrest.03)` %in% c("Not Attempted-Considered\r\nFutile", "Not Attempted-DNR Orders"), na.rm = TRUE),
    other = sum(`Resuscitation Attempted by EMS LIST (eArrest.03)` %in% c("Not Applicable", "Not Recorded"), na.rm = TRUE)
  )

# Print summary of interventions
print(interventions_summary)

# Prepare data for visualization
interventions_data <- tibble(
  Intervention = c("CPR Provided", "Defibrillator Used", "Not Attempted", "Other"),
  Count = c(interventions_summary$cpr_provided, interventions_summary$defibrillator_used, interventions_summary$not_attempted, interventions_summary$other)
)

# Reorder the Intervention factor based on the Count values
interventions_data <- interventions_data %>%
  mutate(Intervention = reorder(Intervention, -Count))  # Reorder by descending count

# Visualize the summary of care provided
ggplot(interventions_data, aes(x = Intervention, y = Count, fill = Intervention)) +
  geom_bar(stat = "identity") +
  labs(title = "Summary of Interventions for Cardiac Arrest Cases (n=33)", x = "Intervention Type", y = "Count") +
  theme_minimal() +
  scale_fill_manual(values = c("gray", "darkorange", "lightblue", "steelblue"))  # Customize colors
```

# Opportunities for Improvement
## Response Times
```{r}
# Convert the date-time columns to POSIXct if they are not already in that format
combined_data <- combined_data %>%
  mutate(
    `Unit Notified by Dispatch Date Time (eTimes.03)` = as.POSIXct(`Unit Notified by Dispatch Date Time (eTimes.03)`, format = "%Y-%m-%d %H:%M:%S"),
    `Unit Arrived on Scene Date Time (eTimes.06)` = as.POSIXct(`Unit Arrived on Scene Date Time (eTimes.06)`, format = "%Y-%m-%d %H:%M:%S"),
    `Unit Left Scene Date Time (eTimes.09)` = as.POSIXct(`Unit Left Scene Date Time (eTimes.09)`, format = "%Y-%m-%d %H:%M:%S"),
    `Unit Back in Service Date Time (eTimes.12` = as.POSIXct(`Unit Back in Service Date Time (eTimes.12`, format = "%Y-%m-%d %H:%M:%S")
  )

# Filter out negative and unrealistic response times
combined_data_clean <- combined_data %>%
  mutate(
    response_time = as.numeric(difftime(`Unit Arrived on Scene Date Time (eTimes.06)`, `Unit Notified by Dispatch Date Time (eTimes.03)`, units = "mins")),
    on_scene_time = as.numeric(difftime(`Unit Left Scene Date Time (eTimes.09)`, `Unit Arrived on Scene Date Time (eTimes.06)`, units = "mins")),
    back_in_service_time = as.numeric(difftime(`Unit Back in Service Date Time (eTimes.12`, `Unit Left Scene Date Time (eTimes.09)`, units = "mins"))
  ) %>%
  # Filter out rows with negative times or times greater than a realistic threshold (e.g., 24 hours for response time)
  filter(
    response_time >= 0 & response_time <= 1440,  # Assuming response times should be within 0 to 24 hours
    on_scene_time >= 0 & on_scene_time <= 1440,  # Assuming on-scene times should be within 0 to 24 hours
    back_in_service_time >= 0 & back_in_service_time <= 1440  # Assuming back-in-service times should be within 0 to 24 hours
  )

# Check the summary of cleaned data
summary(combined_data_clean[c("response_time", "on_scene_time", "back_in_service_time")])

# Recalculate summary statistics for response times
response_time_summary_clean <- combined_data_clean %>%
  filter(`Cardiac Arrest During EMS Event (eArrest.01)` %in% c(
    "Yes, Prior to Any EMS Arrival\r\n(includes Transport EMS &\r\nMedical First Responders)",
    "Yes, After Any EMS Arrival\r\n(includes Transport EMS &\r\nMedical First Responders)"
  )) %>%
  summarise(
    count = n(),
    average_response_time = mean(response_time, na.rm = TRUE),
    median_response_time = median(response_time, na.rm = TRUE),
    min_response_time = min(response_time, na.rm = TRUE),
    max_response_time = max(response_time, na.rm = TRUE)
  )

# Print cleaned summary statistics for response times
print(response_time_summary_clean)

# Filter the data for cardiac arrest events
filtered_data <- combined_data_clean %>%
  filter(`Cardiac Arrest During EMS Event (eArrest.01)` %in% c(
    "Yes, Prior to Any EMS Arrival\r\n(includes Transport EMS &\r\nMedical First Responders)",
    "Yes, After Any EMS Arrival\r\n(includes Transport EMS &\r\nMedical First Responders)"
  ))

# Plot the response times for filtered data with x-axis limited to 200 minutes
ggplot(filtered_data, aes(x = response_time)) +
  geom_histogram(bins = 30, fill = "blue", alpha = 0.7) +
  labs(
    title = "Distribution of Response Times for Cardiac Arrest Events (n=9)",
    x = "Response Time (Minutes)",
    y = "Frequency"
  ) +
  xlim(0, 20) +  # Limit the x-axis to a maximum of 20
  theme_minimal()
```


# Data Quality Concerns and Limitations
## Data Completeness and Identifying Missing Data
```{r}
# Checking data completeness and identifying missing data
missing_data_summary <- sapply(combined_data, function(x) sum(is.na(x)))
print(missing_data_summary)

# Identify columns with more than a certain percentage of missing values
missing_data_percentage <- sapply(combined_data, function(x) mean(is.na(x))) * 100
columns_with_high_missing_data <- names(missing_data_percentage[missing_data_percentage > 20])  # Arbitrary 20% threshold
print(columns_with_high_missing_data)
```

## Identifying potential biases
```{r}
# Check if certain types of cardiac arrest etiologies are reported more frequently
bias_check_etiology <- combined_data %>%
  filter(`Cardiac Arrest During EMS Event (eArrest.01)` %in% c(
    "Yes, Prior to Any EMS Arrival\r\n(includes Transport EMS &\r\nMedical First Responders)",
    "Yes, After Any EMS Arrival\r\n(includes Transport EMS &\r\nMedical First Responders)"
  )) %>%
  group_by(`Cardiac Arrest Etiology (eArrest.02)`) %>%
  summarise(count = n()) %>%
  arrange(desc(count))

# Print summary to check for potential biases in etiology reporting
print(bias_check_etiology)

# Check if resuscitation is attempted more frequently under specific conditions
bias_check_resuscitation <- combined_data %>%
  filter(`Cardiac Arrest During EMS Event (eArrest.01)` %in% c(
    "Yes, Prior to Any EMS Arrival\r\n(includes Transport EMS &\r\nMedical First Responders)",
    "Yes, After Any EMS Arrival\r\n(includes Transport EMS &\r\nMedical First Responders)"
  )) %>%
  group_by(`Resuscitation Attempted by EMS LIST (eArrest.03)`) %>%
  summarise(count = n()) %>%
  arrange(desc(count))

# Print summary to check for potential biases in resuscitation attempts
print(bias_check_resuscitation)
```

## Temporal limitations
```{r}
combined_data <- combined_data %>%
  mutate(`Unit Notified by Dispatch Date Time (eTimes.03)` = as.POSIXct(`Unit Notified by Dispatch Date Time (eTimes.03)`, format = "%Y-%m-%d %H:%M:%S"))

# Check if the conversion worked
str(combined_data$`Unit Notified by Dispatch Date Time (eTimes.03)`)

# Analyze data distribution over time
# Group by month and count the number of incidents
combined_data %>%
  mutate(month = floor_date(`Unit Notified by Dispatch Date Time (eTimes.03)`, "month")) %>%
  group_by(month) %>%
  summarise(count = n()) %>%
  ggplot(aes(x = month, y = count)) +
  geom_line() +
  labs(title = "Incidents Over Time", x = "Month", y = "Number of Incidents") +
  theme_minimal()
```
